\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-08 chapters/abstract.tex
Proofread

\chapter*{Abstract}
Individuals with severe motor impairments, such as those with \ac{lis}
or eye-motor dysfunction, face substantial challenges when using traditional
gaze-dependent brain-computer interfaces \acp{bci}.
These systems require users to direct their gaze toward specific targets, a
task that becomes unfeasible for individuals with limited or no eye control.
This work addresses this limitation by developing gaze-independent
\ac{bci} methods, focusing on improving covert \ac{vsa}, where users can direct
their attention to a target without corresponding eye movements.
A key contribution of this work is the compensation for ERP latency jitter, a
variability that negatively impacts decoding performance in covert \ac{vsa}.
By handling this jitter, the proposed method enhances communication accuracy
without requiring gaze control, making \acp{bci} more usable for individuals with
motor impairments.

>>>>

\chapter*{Abstract}

Individuals with severe motor impairments, such as those with \ac{lis}
or eye-motor dysfunction, face substantial challenges when using traditional
gaze-dependent brain-computer interfaces \acp{bci}.
These systems require users to direct their gaze toward specific targets, a
task that becomes unfeasible for individuals with limited or no eye control.
This work addresses this limitation by developing gaze-independent
\ac{bci} methods, focusing on improving covert \ac{vsa}, where users can direct
their attention to a target without corresponding eye movements.
A key contribution of this work is the compensation for ERP latency jitter, a
variability that negatively impacts decoding performance in covert \ac{vsa}.
By handling this jitter, the proposed method enhances communication accuracy
without requiring gaze control, making \acp{bci} more usable for individuals with
motor impairments.

### 2024-10-08 chapters/bci.tex
Proofread

\todo{comments liesel paradigms}
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\todo{search for ??}
\todo{acro usage in chapters}
\todo{are textcites used when necessary?}

\chapter{Brain-computer interfaces}
\label{sec:bci}
People with \ac{sspi} due to neurodegenerative disease, stroke, traumatic brain or spine
injury become entirely reliant on caregivers or family members in their day-to-day
routine.
Wouldn't it be great if we could help them regain some quality of life by making them
more independent?

@@

Simply put, a \ac{bci} records the user's brain activity, extracts some relevant output
from this brain activity, and couples this output to a function of a device as
illustrated in \autoref{fig:bci/loop}.
Optionally, the user can then observe the action of the device and adapt their behavior
or brain activity accordingly, closing the loop\footnote{\textcite{BCISociety2024} has
recently formalized this into the following definition: \it``A brain-computer interface
is a system that measures brain activity and converts it in (nearly) real-time into
functionally useful outputs to replace, restore, enhance, supplement, and/or improve the
natural outputs of the brain, thereby changing the ongoing interactions between the
brain and its external or internal environments.
It may additionally modify brain activity using targeted delivery of stimuli to create
functionally useful inputs to the brain.''}.

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
}{%
  The \acs{bci} loop.
}{%
  The \ac{bci} loop.
  The user interacts with the \ac{bci} via a specific paradigm, in this case involving
  visual stimulation.
  Electrical neural signals are recorded, and neurophysiological features related to the
  paradigm are extracted.
  Using machine learning, a decision can then be made based on these features, which can
  be presented back to the user.
  In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{%
  fig:bci/loop
}

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of what is
going on in the brain.
Multiple signals can be recorded from the brain, including blood flow and magnetic or
electrical activity.
For several reasons laid out in \autoref{sec:bci-recording}, the electrical modality is
especially interesting.

The recorded brain activity forms only one part of the interaction scheme presented in
\autoref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain activity
to try to extract the desired output.
Instead, it would help greatly if we knew which specific activity to look for.
A \ac{bci} operates according to a specified \emph{paradigm}, which defines the method
by which interaction is conducted.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research to couple
specific activity in the brain to specific information or actions we want to convey via
the \ac{bci}.
The brain signal now \emph{modulates} this information.
In turn, the detection of this signal can be coupled to an action, like typing the
intended letter A or B in \autoref{fig:bci/loop}.
The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the exact
task performed by the user, and is often closely linked to the manner of feedback in the
case of a closed-loop system.
As shown in \autoref{sec:bci/paradigms}, there is a wide variety of \ac{bci} paradigms
based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific brain
signatures related to our paradigm need to be identified within the recorded activity.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks,' all of which generate their own brain activity.
On the other hand, the activity of interest often originates from a specific region or
network within the brain.
This activity might not easily be discerned from all other brain and environmental
activity.
Compare this to trying to listen to a single speaker at a noisy party where everyone is
shouting over each other.
Some obvious sources of interference can be easily filtered out by \emph{preprocessing}
(see \autoref{sec:bci/preprocessing}) and by choosing the correct representation of the
signal (i.e., \emph{feature extraction}), but the problem of retrieving or
\emph{decoding} only the relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm can be retrieved
from it by the decoder and interpreted in function of the desired \ac{bci} output.
\autoref{sec:bci/decoding} takes a closer look at some common decoding techniques.

@@

The \emph{invasiveness} of the technology forms an important concept in determining its
suitability for a given application and patient.
As illustrated in \autoref{fig:bci/recording}, invasive electrodes can often measure a
very specific brain region and can result in better signal quality, at the cost of the
risks involved with brain surgery.

@@

The \ac{bci} paradigm~\cite{Xu2021,Neeling2019} is the key to translating brain activity
into useful output actions, since it defines which neural responses will be elicited and
captured as features.
A paradigm can be loosely seen as \emph{a way of interacting with the \ac{bci}}.
According to \textcite{Zander2011}, they can be arranged by their reliance on external
stimulation and the engagement of the user as shown in \autoref{fig:bci/paradigms}.

@@

Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can be
calculated as
\begin{equation}
	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of different options
that can be selected per trial\footnote{Given that selections are independent of each
other. This formula requires some adaptations in the case of sequential or hierarchical
selection.}, $P$, the selection accuracy of the decoder, and $Q$, the number of trials
per second.
The parameters $N$, $P$, and $Q$ of this formula give us some insight into the building
blocks of a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a broad range of
informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for decoding,
and
\item increase $Q$ by selecting a paradigm that allows fast stimulation or responses.
\end{enumerate*}

@@

\begin{figure}[t]
  \begin{subfigure}[t]{.45\textwidth}
  \includegraphics[width=\textwidth, height=\textwidth]{figures/bci/illustration}
    \caption{Stimuli are flashed while the \ac{eeg} is recorded}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/p300.tikz.tex}
    \caption{The P300 matrix speller interface. Entire rows or columns flash at once.}
  \end{subfigure}

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/erp.tikz.tex}
    \caption{The idealized \ac{erp} evoked by an attended stimulation in a single
    channel.}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth, height=0.6180\textwidth]{example-image-b}
  \end{subfigure}
\end{figure}
\todo{full spatiotemporal erp response}

@@

\subsection{Preprocessing \& feature extraction}
For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5Hz and
16Hz.
Re-referencing can be done to average mastoids to highlight the contribution of the P3
component.
In addition to the steps outlined in \autoref{sec:bci/preprocessing}, the signal is
usually cut into \emph{epochs} for \ac{erp} analysis.
These are time windows surrounding the stimulus event.
They usually include some baseline interval before the stimulus, which can be used to
normalize the response.

@@

With this distance defined, the epochs can either be directly classified on the
Riemannian manifold using a minimum distance to the mean classifier, or they can be
projected to the tangent space.
This is a Euclidean space that locally approximates the Riemannian distance, which
allows us to apply regular machine learning methods like an SVM or \ac{lda}.
The tangent space is obtained\todo{tangent space}.
To enhance classifications of \acp{erp} which rely on time-domain information, which
would otherwise be lost in the spatial representation of the extracted feature, the
class averages can be concatenated as additional channels to the epochs, embedding the
temporal structure in the feature~\cite{Barachant2014}.
To improve performance even further, the \acp{erp} can be decomposed by a method like
XDAWN before calculating the extended covariances~\cite{Li2020}.

@@

Furthermore, the \ac{eeg}-based visual event-related potential (\ac{erp})-based oddball
interface is an effective and proven method~\cite{Wolpaw2018,Severens2020}.
Targets are shown in short flashes on a computer screen that evoke an \ac{erp} when
observed, which can be detected in the \ac{eeg} signal.
The \ac{erp} consists of multiple components, some of which are modulated by perception
and the attention of the user.
Decoding this modulation allows for information transfer controlled by the user's brain
activity, as sketched in \autoref{fig:bci/loop}.

@@

Different degrees of eye motor impairment can render visual \ac{bci}s uncomfortable or
outright impossible to use.
When operating a visual \ac{bci}, the usual rapid series of forced saccades followed by
fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest position of
their choice while operating the \ac{bci}.
This points to the need for gaze-independent solutions, which we will explore further in
\autoref{sec:gaze-independence} and the rest of this thesis.

>>>>

\chapter{Brain-computer interfaces}
\label{sec:bci}
\todo{search for ??}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todo{fix all latex warnings to make sure no refs are missing}
\todo{uniform ref or cref}
\todo{consistently terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todo{uniform vsa condition vsa setting}
\todo{are all acronyms defined/Acro corectly used in chapters?}
\todo{acronyms written in full every new chapter and in captions?}
\todo{are textcites used when necessary?}
\todo{fix lacheck/chktex notifications}
\todo{footnot skips a page in wcble chap}

People with \ac{sspi} due to neurodegenerative disease, stroke, traumatic brain or spine
injury become entirely reliant on caregivers or family members in their day-to-day
routine.
Wouldn't it be great if we could help them regain some quality of life by making them
more independent?

@@

Simply put, a \ac{bci} records the user's brain activity, extracts some relevant output
from this brain activity, and couples this output to a function of a device as
illustrated in \cref{fig:bci/loop}.
Optionally, the user can then observe the action of the device and adapt their behavior
or brain activity accordingly, closing the loop\footnote{\textcite{BCISociety2024} has
recently formalized this into the following definition: \it``A brain-computer interface
is a system that measures brain activity and converts it in (nearly) real-time into
functionally useful outputs to replace, restore, enhance, supplement, and/or improve the
natural outputs of the brain, thereby changing the ongoing interactions between the
brain and its external or internal environments.
It may additionally modify brain activity using targeted delivery of stimuli to create
functionally useful inputs to the brain.''}.

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
}{%
  The \acs{bci} loop.
}{%
  The \ac{bci} loop.
  The user interacts with the \ac{bci} via a specific paradigm, in this case involving
  visual stimulation.
  Electrical neural signals are recorded, and neurophysiological features related to the
  paradigm are extracted.
  Using machine learning, a decision can then be made based on these features, which can
  be presented back to the user.
  In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{fig:bci/loop}

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of what is
going on in the brain.
Multiple signals can be recorded from the brain, including blood flow and magnetic or
electrical activity.
For several reasons laid out in \cref{sec:bci-recording}, the electrical modality is
especially interesting.

The recorded brain activity forms only one part of the interaction scheme presented in
\cref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain activity
to try to extract the desired output.
Instead, it would help greatly if we knew which specific activity to look for.
A \ac{bci} operates according to a specified \emph{paradigm}, which defines the method
by which interaction is conducted.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research to couple
specific activity in the brain to specific information or actions we want to convey via
the \ac{bci}.
The brain signal now \emph{modulates} this information.
In turn, the detection of this signal can be coupled to an action, like typing the
intended letter A or B in \cref{fig:bci/loop}.
The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the exact
task performed by the user, and is often closely linked to the manner of feedback in the
case of a closed-loop system.
As shown in \cref{sec:bci/paradigms}, there is a wide variety of \ac{bci} paradigms
based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific brain
signatures related to our paradigm need to be identified within the recorded activity.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks,' all of which generate their own brain activity.
On the other hand, the activity of interest often originates from a specific region or
network within the brain.
This activity might not easily be discerned from all other brain and environmental
activity.
Compare this to trying to listen to a single speaker at a noisy party where everyone is
shouting over each other.
Some obvious sources of interference can be easily filtered out by \emph{preprocessing}
(see \cref{sec:bci/preprocessing}) and by choosing the correct representation of the
signal (i.e., \emph{feature extraction}), but the problem of retrieving or
\emph{decoding} only the relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm can be retrieved
from it by the decoder and interpreted in function of the desired \ac{bci} output.
\Cref{sec:bci/decoding} takes a closer look at some common decoding techniques.

@@

The \emph{invasiveness} of the technology forms an important concept in determining its
suitability for a given application and patient.
As illustrated in \cref{fig:bci/recording}, invasive electrodes can often measure a
very specific brain region and can result in better signal quality, at the cost of the
risks involved with brain surgery.

@@

The \ac{bci} paradigm~\cite{Xu2021,Neeling2019} is the key to translating brain activity
into useful output actions, since it defines which neural responses will be elicited and
captured as features.
A paradigm can be loosely seen as \emph{a way of interacting with the \ac{bci}}.
According to \textcite{Zander2011}, they can be arranged by their reliance on external
stimulation and the engagement of the user as shown in \cref{fig:bci/paradigms}.

@@

Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can be
calculated as
\begin{equation}
  \text{\acs{itr}} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of different options
that can be selected per trial\footnote{Given that selections are independent of each
other. This formula requires some adaptations in the case of sequential or hierarchical
selection.}, $P$, the selection accuracy of the decoder, and $Q$, the number of trials
per second.
The parameters $N$, $P$, and $Q$ of this formula give us some insight into the building
blocks of a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a broad range of
informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for decoding,
and
\item increase $Q$ by selecting a paradigm that allows fast stimulation or responses.
\end{enumerate*}

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[t]{.45\textwidth}
  \includegraphics[width=\textwidth, height=\textwidth]{figures/bci/illustration}
    \caption{Stimuli are flashed while the \acs{eeg} is recorded.}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/p300.tikz.tex}
    \caption{The P300 matrix speller interface. Entire rows or columns flash at once.}
  \end{subfigure}

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/erp.tikz.tex}
    \caption{The idealized \ac{erp} evoked by an attended stimulation in a single
    channel.}
  \end{subfigure}%
\end{figure}

@@

\subsection{Preprocessing \& feature extraction}
For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5Hz and
16Hz.
Re-referencing can be done to average mastoids to highlight the contribution of the P3
component.
In addition to the steps outlined in \cref{sec:bci/preprocessing}, the signal is
usually cut into \emph{epochs} for \ac{erp} analysis.
These are time windows surrounding the stimulus event.
They usually include some baseline interval before the stimulus, which can be used to
normalize the response.

@@

With this distance defined, the epochs can either be directly classified on the
Riemannian manifold using a minimum distance to the mean classifier, or they can be
projected to the tangent space.
This is a Euclidean space that locally approximates the Riemannian distance, which
allows us to apply regular machine learning methods like an SVM or \ac{lda}.
%The tangent space is obtained\todo{tangent space}.
To enhance classifications of \acp{erp} which rely on time-domain information, which
would otherwise be lost in the spatial representation of the extracted feature, the
class averages can be concatenated as additional channels to the epochs, embedding the
temporal structure in the feature~\cite{Barachant2014}.
To improve performance even further, the \acp{erp} can be decomposed by a method like
XDAWN before calculating the extended covariances~\cite{Li2020}.

@@

Furthermore, the \ac{eeg}-based visual event-related potential (\ac{erp})-based oddball
interface is an effective and proven method~\cite{Wolpaw2018,Severens2014}.
Targets are shown in short flashes on a computer screen that evoke an \ac{erp} when
observed, which can be detected in the \ac{eeg} signal.
The \ac{erp} consists of multiple components, some of which are modulated by perception
and the attention of the user.
Decoding this modulation allows for information transfer controlled by the user's brain
activity, as sketched in \cref{fig:bci/loop}.

@@

Different degrees of eye motor impairment can render visual \ac{bci}s uncomfortable or
outright impossible to use.
When operating a visual \ac{bci}, the usual rapid series of forced saccades followed by
fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest position of
their choice while operating the \ac{bci}.
This points to the need for gaze-independent solutions, which we will explore further in
\cref{sec:gaze-independence} and the rest of this thesis.
